The relative roles of genetic, ecological and environmental factors in species radiations

### 2018-02-20 paper/draft.tex
restructure flow of intro

@@

%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}
Still, the proposed mechanisms so far include only ecological or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
while genetic properties are hardly recognized in the island context.
At the same time, an increasing number of studies investigates genetic processes and
their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein).

In a recent review \cite{leidinger2017biodiversity}, we analyzed a powerful alternative to field experiments in island biogeography: mechanistic (/simulation) models.
Although many of the reviewed models integrate multiple processes at a time, unfortunately, most models failed to include more explicit genetic or micro evolution processes.
Instead, some models used concepts such as ``protracted speciaton'' \cite{rosindellXXX},
where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms can have important ramifications for speciation/radiation and species richness.

%%TODO: here come a list of processes with examples
%% - genetic processes in general
%% - isolation processes

For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013) % cite properly!

Another shortfall of the reviewed models was the lack of integrating ecological niches and adaption.

Besides adaptive radiations(/evolution), changes in or different measures of geographic isolation (landscape complexity) can also give rise to non-adaptive radiations.

In the following, we present a mechanistic model that remedies both of the adressed shortcomings
by implementing genetic processes on a niche-explicit background
using an island biogeography framework.%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
%% and what trait combinations constitute successful colonizer and radiating species, respectively.
We will adress the questions of what relative impact does (a) genetic linkage and
(b) reproductive isolation have on emerging species numbers and diversity,
and (c) which genetic and functional trait characteristics distinguish successful colonizer,
strong competitor and diversifying lineages.

\section{Material and Methods}
We use an individual based metacommunity model written in julia \cite{JULIALANG}.
Each individual is characterized by a set of genetically controlled traits and is submitted to processes of
establishment, competition, growth, (density independent) mortality, reproduction and dispersal.
Each of which process is affected by the individuals' trait characteristics.
The simulation arena consists of a larger landmass (the mainland) holding an initial random species pool and a smaller, initially unpopulated island some distance apart.
During the course of the simulation, individuals may disperse from the mainland to the island, thus colonizing it.
The complete state of the simulation is recorded every 1000 time steps for analysis.

>>>>

In a recent review %\cite{leidinger2017biodiversity}
(Leidinger \& Sarmento Cabral 2017), we analyzed a powerful alternative to field experiments in island biogeography: mechanistic (/simulation) models.
Although many of the reviewed models integrate multiple processes at a time, unfortunately, most models failed to include more explicit genetic or micro evolution processes.
Instead, some models used concepts such as ``protracted speciaton'' (Rosindell \& Harmon 2011) %\cite{rosindellXXX},
where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this (cf.\ Silvertown 2004, but see Herben et al.\ 2005) and genetic mechanisms can have important ramifications for speciation/radiation and species richness.

@@

%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}
Still, the proposed mechanisms so far include only ecological or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
while genetic properties are hardly recognized in the island context.
At the same time, an increasing number of studies investigates genetic processes and
their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein).
%%TODO: here come a list of processes with examples
%% - genetic processes in general
%% - isolation processes
Besides arthropods (Howthorne \& Via 2001), these processes can be observed in plants as well.
For example, the rapid switching of pollinator syndromes in species of the genus \textit{Petunia} %use package for this!
seems to be a consequence of the tight genetic linkage between loci for pollination syndrome traits (Hermann et al.\ 2013). % cite properly!
Considering these phenomena, we expect speciation rates to be highest at intermediate levels of gene linkage --- low
levels leads to dissolving phenotype populations, whereas high levels hinder rapid adaption.

%Another shortfall of the reviewed models was the lack of integrating ecological niches and adaption.

%Besides adaptive radiations(/evolution), changes in or different measures of geographic isolation (landscape complexity) can also give rise to non-adaptive radiations.

To address this hypothesis, we present a mechanistic model that tries to remedy above mentioned shortcomings
by implementing genetic processes on a niche-explicit background
using an island biogeography framework.%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
%% and what trait combinations constitute successful colonizer and radiating species, respectively.
We will adress the questions of what relative impact does (a) genetic linkage and
(b) reproductive isolation have on emerging species numbers and diversity,
and (c) which genetic and functional trait characteristics distinguish successful colonizer,
strong competitor and diversifying lineages.

\section{Material and Methods}
We use an individual based metacommunity model written in julia %\cite{JULIALANG}.
Each individual is characterized by a set of genetically controlled traits and is submitted to processes of
establishment, competition, growth, (density independent) mortality, reproduction and dispersal.
Each of which process is affected by the individuals' trait characteristics.
The simulation arena consists of a larger landmass (the mainland) holding an initial random species pool and a smaller, initially unpopulated island some distance apart.
During the course of the simulation, individuals may disperse from the mainland to the island, thus colonizing it.
The complete state of the simulation is recorded every 1000 time steps for analysis.
